sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-03-11 doc/paper-jss/paper-jss.Rmd
Formatting #13

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

```{r main-setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "> ")
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

@@

```{r pcod-utms-eval, echo=FALSE, eval=TRUE}
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
```

```{r pcod-utms-echo, echo=TRUE, eval=FALSE}
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
#> Detected UTM zone 9N; CRS = 32609.
#> Visit https://epsg.io/32609 to verify.
```

@@

\caption{Comparison of functionality between several \proglang{R} packages that can fit geostatistical GLMMs.}
\centering
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllllll}
\toprule
 & \pkg{sdmTMB} & \pkg{VAST} & \proglang{R}-\pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
\midrule
Time-varying coefficients & \checkmark & --$^1$ & \checkmark & \checkmark & \checkmark & --\\
Spatially varying coefficients (SVC) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
GAMs$^2$ & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Threshold covariates & \checkmark & -- & \checkmark$^3$ & -- & -- & --\\
Offsets & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
\addlinespace
Spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark$^4$ & --\\
Spatial + spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Anisotropy & \checkmark & \checkmark & -- & \checkmark & -- & --\\
Correlation barriers & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Separate range parameters for fields & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Share range parameters across fields & \checkmark & \checkmark & \checkmark & -- & -- & --\\
SPDE-based & \checkmark & \checkmark & \checkmark & \checkmark$^5$ & --$^6$ & \checkmark\\
\addlinespace
NB1 distribution & \checkmark & -- & \checkmark & \checkmark & -- & \checkmark\\
NB2 distribution & \checkmark & \checkmark$^7$ & \checkmark & \checkmark & -- & \checkmark\\
Zero-truncated distributions & \checkmark & -- & \checkmark & -- & -- & \checkmark\\
Zero-inflated distributions & \checkmark & \checkmark & \checkmark & -- & -- & \checkmark\\
Tweedie distribution & \checkmark & \checkmark & \checkmark & \checkmark$^8$ & -- & --\\
Student-t distribution & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Censored Poisson distribution & \checkmark & -- & \checkmark & -- & -- & --\\
log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
Poisson-link delta model & \checkmark & \checkmark & \checkmark & -- & -- & --\\
Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
Maximum/marginal likelihood & \checkmark & \checkmark & -- & \checkmark & -- & --\\
\addlinespace
Bayesian/optionally Bayesian & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
Priors/penalties & \checkmark & -- & \checkmark & -- & \checkmark & --\\
Matern PC priors & \checkmark & -- & \checkmark & -- & -- & --\\
\addlinespace
Spatial (or spatial dynamic) factor analysis & -- & \checkmark & -- & -- & -- & --\\
Empirical Orthogonal Function (EOF) analysis & -- & \checkmark & -- & -- & -- & --\\
Built-in area-weighted index standardization & \checkmark & \checkmark & -- & -- & -- & --\\
Built-in cross-validation & \checkmark & -- & -- & -- & -- & --\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: }
\item $^1$Technically possible but non-trivial. $^2$Penalized smoother GAMs that determine `wiggliness'. $^3$\pkg{inlabru} but not \proglang{R}-\pkg{INLA}. $^4$Spatiotemporal fields as random walk only. $^5$SPDE approach as in \citet{miller2019}. $^6$Does have predictive process knots. $^7$Zero-inflated NB2 only. $^8$Tweedie power parameter fixed for \texttt{mgcv::gamm()}. $^9$Possible as log-linked Poisson GLMM with aggregated data. $^{10}$Hurdle models possible by fitting components separately.
\end{tablenotes}
\end{threeparttable}
\label{tab:functionality}
\end{table}

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \usepackage{gensymb}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \newunicodechar{˚}{\degree}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

```{r main-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  # prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "+ ")
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
```

@@

```{r pcod-utms-eval, echo=TRUE, eval=TRUE}
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
```

```
#> Detected UTM zone 9N; CRS = 32609.
#> Visit https://epsg.io/32609 to verify.
```

@@

\centering
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllllll}
\toprule
 & \pkg{sdmTMB} & \pkg{VAST} & \proglang{R}-\pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
\midrule
Time-varying coefficients & \checkmark & --$^1$ & \checkmark & \checkmark & \checkmark & --\\
Spatially varying coefficients (SVC) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
GAMs$^2$ & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Threshold covariates & \checkmark & -- & \checkmark$^3$ & -- & -- & --\\
Offsets & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
\addlinespace
Spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark$^4$ & --\\
Spatial + spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Anisotropy & \checkmark & \checkmark & -- & \checkmark & -- & --\\
Correlation barriers & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Separate range parameters for fields & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Share range parameters across fields & \checkmark & \checkmark & \checkmark & -- & -- & --\\
SPDE-based & \checkmark & \checkmark & \checkmark & \checkmark$^5$ & --$^6$ & \checkmark\\
\addlinespace
NB1 distribution & \checkmark & -- & \checkmark & \checkmark & -- & \checkmark\\
NB2 distribution & \checkmark & \checkmark$^7$ & \checkmark & \checkmark & -- & \checkmark\\
Zero-truncated distributions & \checkmark & -- & \checkmark & -- & -- & \checkmark\\
Zero-inflated distributions & \checkmark & \checkmark & \checkmark & -- & -- & \checkmark\\
Tweedie distribution & \checkmark & \checkmark & \checkmark & \checkmark$^8$ & -- & --\\
Student-t distribution & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Censored Poisson distribution & \checkmark & -- & \checkmark & -- & -- & --\\
log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
Poisson-link delta model & \checkmark & \checkmark & \checkmark & -- & -- & --\\
Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
Maximum/marginal likelihood & \checkmark & \checkmark & -- & \checkmark & -- & --\\
\addlinespace
Bayesian/optionally Bayesian & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
Priors/penalties & \checkmark & -- & \checkmark & -- & \checkmark & --\\
Matern PC priors & \checkmark & -- & \checkmark & -- & -- & --\\
\addlinespace
Spatial (or spatial dynamic) factor analysis & -- & \checkmark & -- & -- & -- & --\\
Empirical Orthogonal Function (EOF) analysis & -- & \checkmark & -- & -- & -- & --\\
Built-in area-weighted index standardization & \checkmark & \checkmark & -- & -- & -- & --\\
Built-in cross-validation & \checkmark & -- & -- & -- & -- & --\\
\bottomrule
\end{tabular}
\end{threeparttable}
\label{tab:functionality}

\caption{Comparison of functionality between several \proglang{R} packages that can fit geostatistical GLMMs.
Notes: 
$^1$Technically possible but non-trivial.
$^2$Penalized smoother GAMs that determine `wiggliness'.
$^3$\pkg{inlabru} but not \proglang{R}-\pkg{INLA}.
$^4$Spatiotemporal fields as random walk only.
$^5$SPDE approach as in \citet{miller2019}.
$^6$Does have predictive process knots.
$^7$Zero-inflated NB2 only.
$^8$Tweedie power parameter fixed for \texttt{mgcv::gamm()}.
$^9$Possible as log-linked Poisson GLMM with aggregated data.
$^{10}$Hurdle models possible by fitting components separately.
}
\end{table}
